---
title: "Damei"
author: "Allen"
date: "27/05/2017"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=TRUE, message=FALSE, warning=FALSE}
setwd("/Users/Allen/Desktop")
library(xlsx)
Damei<-read.xlsx2("damei.xlsx",sheetIndex = 1,header = TRUE)
Damei<-Damei[Damei$联系人 %in% c("我","yupeiran"),]
Damei<-droplevels(Damei)
str(Damei)
```

Deal with time data, remove records before 11th May 2017, 
create weekdays, time range，date， time
change 联系人 to 1&2
```{r, echo=TRUE, message=FALSE, warning=FALSE}
Damei$时间<-as.POSIXlt(strptime(Damei$时间,"%d/%m/%Y %H:%M"))
Damei<-Damei[Damei$时间>as.POSIXlt("2017-05-11 23:55:00"),]
Damei$day<-weekdays(Damei$时间)
Damei$day<-as.factor(Damei$day)
Damei$day <- factor(Damei$day,levels(Damei$day)[c(2,6,7,5,1,3,4)])

Damei$hour<-Damei$时间$hour

Damei$timerange<-Damei$hour
Damei$timerange[Damei$timerange==0 | Damei$timerange==1]<-"0-2"
Damei$timerange[Damei$timerange==2 | Damei$timerange==3]<-"2-4"
Damei$timerange[Damei$timerange==4 | Damei$timerange==5]<-"4-6"
Damei$timerange[Damei$timerange==6 | Damei$timerange==7]<-"6-8"
Damei$timerange[Damei$timerange==8 | Damei$timerange==9]<-"8-10"
Damei$timerange[Damei$timerange==10 | Damei$timerange==11]<-"10-12"
Damei$timerange[Damei$timerange==12 | Damei$timerange==13]<-"12-14"
Damei$timerange[Damei$timerange==14 | Damei$timerange==15]<-"14-16"
Damei$timerange[Damei$timerange==16 | Damei$timerange==17]<-"16-18"
Damei$timerange[Damei$timerange==18 | Damei$timerange==19]<-"18-20"
Damei$timerange[Damei$timerange==20 | Damei$timerange==21]<-"20-22"
Damei$timerange[Damei$timerange==22 | Damei$timerange==23]<-"22-24"
Damei$timerange<-as.factor(Damei$timerange)
Damei$timerange<-factor(Damei$timerange, levels(Damei$timerange)[c(1,9,10,2:8)])

Damei$time<-strftime(Damei$时间,format = "%H:%M:%S")

Damei$uniqueday<-Damei$时间$mday
Damei$uniqueday<-as.factor(Damei$uniqueday)

levels(Damei$联系人)<-c(2,1)
Damei$联系人[Damei$联系人=="我"]<-1
Damei$联系人[Damei$联系人=="yupeiran"]<-2
```

create Damei_only dataframe
```{r, echo= TRUE, message=TRUE, warning=FALSE}

DameiO<-Damei[Damei$联系人==2,]
DameiO<-droplevels(DameiO)
DameiO<-DameiO[,-c(2,3)]
```

## Including Plots
histogram
```{r, echo= TRUE}
library(ggplot2)
ggplot(DameiO,aes(x=as.numeric(uniqueday)))+
  geom_histogram(colour="lightblue")+ggtitle("messages by day")
```

overview
```{r}
Dameiplot <- ggplot(DameiO, aes(x = uniqueday, y = timerange))
Dameiplot + geom_jitter(alpha = 0.5,
aes(color = uniqueday),
position = position_jitter(width = 0.2)) + labs(y = "time",x="date") +
ggtitle("Messages by day") + theme(plot.title = element_text(hjust = 0.5))

```

get alternate dataframe. 
select pairs and remove extra rows. the pattern is now 121212...(the last message of allen and first message of Damei)
```{r, echo= TRUE}
Damei$联系人<-as.factor(Damei$联系人)
Alternate<-data.frame(联系人=factor(),stringsAsFactors = FALSE)
number=1
for (i in 1:(nrow(Damei)-1)){
  if (Damei$联系人[i]=="1" &Damei$联系人[i+1]=="2"){
    z<-rbind(Damei[i,],Damei[i+1,])
    rownames(z)<-c(number,number+1)
    Alternate<-rbind(Alternate,z)
    number<-number+2
  }
}
```

find the replying time for all my message
```{r, echo= TRUE}

for (i in 1:(nrow(Alternate)-1)){
    Alternate$replytime[i]<-difftime(Alternate$时间[i+1], Alternate$时间[i],units = "mins")
  }

```

find average replying time
```{r, echo=TRUE}
mean(Alternate[Alternate$联系人==1,]$replytime)
```

identify converstaions
if next message is within 4 minutes, select these messages. converstaion:121212. conversationD:22222
```{r, echo= TRUE}
Alternate$conversation<-NA
Alternate$conversation[1]<-1
Alternate$conversation[2]<-1
conversation=1
for (i in 4:(nrow(Alternate)-1)){
   if (Alternate$联系人[i]==2){
     if (difftime(Alternate$时间[i+2], Alternate$时间[i],units = "mins")<4){
         if (difftime(Alternate$时间[i], Alternate$时间[i-2],units = "mins")<4){
         Alternate$conversation[i-1]<-conversation
         Alternate$conversation[i]<-conversation
         Alternate$conversation[i+1]<-conversation
         Alternate$conversation[i+2]<-conversation
         }else{
        conversation<-conversation+1
        Alternate$conversation[i-1]<-conversation
        Alternate$conversation[i]<-conversation
        Alternate$conversation[i+1]<-conversation
        Alternate$conversation[i+2]<-conversation
         }
       i=i+4
    }
  }
}
conversation<-Alternate[!is.na(Alternate$conversation),]
#get Damei only conversation 
conversationD<-conversation[conversation$联系人==2,]
a<-table(conversationD$conversation)
a<-as.data.frame(a)
colnames(a)<-c("conversation","count")
conversationD<-merge(conversationD,a,by = "conversation")
conversationD$conversation<-as.factor(conversationD$conversation)

#keep only the first row of each conversationD
conversationR<-conversationD[!duplicated(conversationD$conversation),]
```

plot conversation against day, boxplot
```{r  , echo= TRUE}
conversationplot<-ggplot(conversationR,aes(x=day,y=count))
conversationplot+geom_boxplot(fill="chartreuse4")+labs(y="Length of conversation")+ggtitle("Conversation by day")+theme(plot.title = element_text(hjust = 0.5))
```

plot conversation against time range, weekday as color
```{r  , echo= TRUE}
conversationplot<-ggplot(conversationR,aes(x=timerange,y=count,color=factor(day)))
conversationplot + geom_jitter(alpha=0.5,aes(color=day),position=position_jitter(width = 0.2)) + labs(y = "Length of conversation") + ggtitle("Conversation by time") +
  theme(plot.title = element_text(hjust = 0.5))
```

plot conversation against day. 
```{r  , echo= TRUE}
conversationplot <- ggplot(conversationR, aes(x = day, y = timerange))
conversationplot + geom_jitter(alpha = 0.5,
aes(color = day),
position = position_jitter(width = 0.2)) + labs(y = "Time") +
ggtitle("Conversation by day") + theme(plot.title = element_text(hjust = 0.5))
```

Now, lets remove conversations from the main dataset and examine how much time Damei takes to reply a message.
```{r, echo= TRUE}
nocon<-Alternate[is.na(Alternate$conversation),]
nocon1<-nocon[nocon$联系人==1,]
mean(nocon1$replytime)
```

remove outlier 
```{r, echo=TRUE}
#assume sleeping time = 8 hours +- 1 hour
nocon2<-nocon1[nocon1$replytime<420,]
```

```{r  , echo= TRUE}
summary(nocon2$replytime)
```

try boxplot
```{r  , echo= TRUE}
noconplot<-ggplot(nocon2,aes(x="",y=replytime))
noconplot+geom_boxplot(fill="chartreuse4")+labs(y="replytime")+ggtitle("Replytime by Shuigui")+theme(plot.title = element_text(hjust = 0.5))
```

classification on conversation/ no conversation 

identify extract conversationA from conversation dataset. 
remove duplicates in conversationA
combine noconversation dataset + conversationA dataset to obtain a combined dataset: AlternateR
```{r  , echo= TRUE}
conversationA<-conversation[conversation$联系人==1,]
a<-table(conversationA$conversation)
a<-as.data.frame(a)
colnames(a)<-c("conversation","count")
conversationA<-merge(conversationA,a,by = "conversation")
conversationA$conversation<-as.factor(conversationA$conversation)

#keep only the first row of each conversationD
conversationAR<-conversationA[!duplicated(conversationA$conversation),]
conversationAR$conversation<-"yes"
conversationAR<-conversationAR[,-13]
nocon1$conversation<-"no"
AlternateR<-rbind(nocon1,conversationAR)
make.names(AlternateR$conversation,unique = FALSE)
AlternateR$conversation<-as.factor(AlternateR$conversation)
levels(AlternateR$conversation)
```

145 samples, we conduct classification, using 类型，time range and day as predictor variables.
```{r  , echo= TRUE}
training = sample(1:145, 100)
train<-AlternateR[training,]
test<-AlternateR[-training,]
Xtrain<-model.matrix(conversation~timerange+day+类型,data=train)
Xtest<-model.matrix(conversation~timerange+day+类型,data=test)
ytrain<-as.factor(train[,"conversation"])
ytest<-as.factor(test[,"conversation"])

#ridge logistic regression
library(glmnet)
lambda.grid = 10^seq( -5, 2,length=100)
ridge.cv = cv.glmnet(Xtrain, ytrain, alpha=0,
                     type.measure = "mse", nfolds = 10,
                     lambda = lambda.grid,family="binomial")

prediction.ridge = predict(ridge.cv, newx = Xtest,
  s = "lambda.min", type="response")

plot.new()
#plot roc
library(pROC)
auc<-roc(ifelse(ytest=="yes",1,0),as.vector(prediction.ridge))
plot(auc,print.auc=TRUE)
```


try boosting
```{r  , echo= TRUE}
library(caret)
library(gbm)
boostytrain<-train[,12]
boostxtrain<-train[,c(4,6,8)]
boostytest<-test[,12]
boostxtest<-test[,c(4,6,8)]

objControl<-trainControl(method = "cv",number=3,returnResamp = "none",summaryFunction = twoClassSummary,classProbs = TRUE)
ObjModel<-train(boostxtrain,boostytrain,method="gbm",trControl = objControl,metric = "ROC",preProcess = c("center","scale"))
summary(ObjModel)
```

```{r  , echo= TRUE}
predictions <- predict(object = ObjModel,boostxtest,type = "prob")
auc<-roc(ifelse(boostytest=="yes",1,0),predictions[[2]])
plot(auc,print.auc=TRUE)
```

Classification for reply time <10 and >10
```{r }
boost2ytrain<-train[,11]
boost2ytrain<-ifelse(boost2ytrain<10,"good","laosy")
boost2ytrain<-as.factor(boost2ytrain)
boost2ytest<-test[,11]
boost2ytest<-ifelse(boost2ytest<10,"good","laosy")
boost2ytest<-as.factor(boost2ytest)
boost2xtrain<-train[,c(4,6,8)]
boost2xtest<-test[,c(4,6,8)]

objControl<-trainControl(method = "cv",number=3,returnResamp = "none",summaryFunction = twoClassSummary,classProbs = TRUE)
ObjModel2<-train(boost2xtrain,boost2ytrain,method="gbm",trControl = objControl,metric = "ROC",preProcess = c("center","scale"))
summary(ObjModel2)
```

```{r  , echo= TRUE}
predictions2 <- predict(object = ObjModel2,boost2xtest,type = "prob")
auc2<-roc(ifelse(boost2ytest=="good",0,1),predictions2[[2]])
plot(auc2,print.auc=TRUE)
```


